Improving the templates for multicolor Ramsey numbers and Schur numbers

### 2022-06-29 article_ramsey/template_improvement.tex
ineqs

Ramsey numbers and Schur numbers are defined in Section \ref{sec:def}. Section \ref{sec:temp} defines the 
\(b\)-templates and describes the corrresponding construction. Section \ref{sec:results} states and proves the main 
results on \(b\)-templates and on the construction. Section \ref{sec:bounds} records the new inequalities and some new 
lower bounds.

@@

Let \(a \in \mathbb{N}^*\), \(b \in [\![0, a - 1]\!]\) and \(r \in \mathbb{N}\). Let 
\((k_c)_{1 \leqslant c \leqslant r} \in {[\![3, + \infty [\![}^r\). Let \(T \in \mathcal{T}_b(k_1, ..., k_r; a)\) and 
denote by \(A_1, ..., A_{r+1}\) the associated partition. Let \(d \in [\![0, a]\!]\). Let \(F\) be a coloring of 
\([\![1, d]\!]\) with \(r +1\) colors and denote by \(B_1, ..., B_{r + 1}\) the associated partition. The coloring 
\(F\) is said to be compatible with the \(b\)-template \(T\) if it satisfies:
\begin{enumerate}[(i)]
\item \(a - b \notin B_{r + 1}\)
\item \(\forall (x,y) \in A_{r+1}^2, (x + y > a + b ~\text{and}~ \pi_{a, b}(x + y) \leqslant d) \implies \pi_{a, b}
	(x + y)  \notin B_{r+1}\)
\item \(\forall (x,y) \in A_{r+1} \times B_{r+1}, x + y \leqslant d \implies x + y  \notin B_{r+1}\)
\item \(\forall c \in [\![1, r]\!], \forall k \in [\![1, k_c-1]\!], \forall x \in \TS_{b, k}(A_c), \forall y \in 
	\mathfrak{I}_{k_c - k - 1}(B_c), \\
	\begin{array}{l}
		\mspace{30mu} 
		\left\{ \text{or} 
		\begin{array}{l}
			\exists (i, j) \in {[\![1, k]\!]}^2, \pi_{a, b}(x_j - x_i) \notin A_c, \\
			\exists (i, j) \in [\![1, k]\!] \times [\![1, k_c - k - 1]\!], \\
			\mspace{30mu}
				\left \{
				\begin{array}{l l}
					(y_j - x_i) \mod a \notin A_c & ~\text{if}~ (y_j - x_i) \mod a > b \\
					(y_j - x_i) \mod a \notin A_c ~\text{and}~ \pi_{a, b}(y_j - x_i) \notin A_c & ~\text{if}~ (y_j - x_i) \mod a 
						\leqslant b \\
				\end{array}
				\right. \\
			\exists (i, j) \in {[\![1, k_c - k - 1]\!]}^2, i < j ~\text{and}~ \pi_{a, b}(y_j - y_i) \notin A_c.
		\end{array}
		\right.
	\end{array}\)
\end{enumerate}
\end{definition}

@@

\begin{corollary}[Inégalités pour les nombres de Ramsey linéaires]
Soit \(r \in \mathbb{N}\) et soit \((k_c)_{1 \leqslant c \leqslant r} \in {[\![3, +\infty[\![}^r\). Soit \(s \in \mathbb{N}\) et soit \((l_c)_{1 \leqslant c \leqslant s} \in {[\![3, +\infty[\![}^s\). Soient \(a \in \mathbb{N}^*\) et \(b \in \mathbb{N}\) tels que \(\mathcal{T}_b(l_1, ..., l_s, t; a) \neq \varnothing\), et on fixe \(T\) un tel \(b\)-template. Soit \(d \in [\![1, a]\!]\) tel qu'il existe un coloriage de \([\![1, d]\!]\) à \(s + 1\) couleurs compatible avec \(T\).
Alors \(L(k_1, ..., k_r, l_1, ..., l_s) - 2 \geqslant a (L(k_1, ..., k_r) - 2) + b + d\).
\end{corollary}

@@

\section{Nouvelles inégalités et bornes inférieures}
\label{sec:bounds}

\subsection{Nombres de Schur}

L'existence d'un 2-S-template de largeur 10 à 3 couleurs non exprimable en tant que S-template fournit une nouvelle inégalité. L'ancienne inégalité était \(S(n + 2) \geqslant 9 \, S(n) + 4\).

\begin{proposition}
	\[\forall n \in \mathbb{N}^*, S(n + 2) \geqslant 10 \, S(n) + 2\]
\end{proposition}

Cette inégalité fournit les nouvelles bornes inférieures suivantes : \(S(8) \geqslant 5\,362\) et \(S(13) \geqslant 2\,038\,282\) (contre respectivement \(5\,286\) dans \cite{RowleyRamsey} et \(2\,011\,290\) dans \cite{schurboyz}). Plus généralement, cette inégalité améliore les bornes inférieures des \(S(5 k + 3), \forall k \in \mathbb{N}^*\).

\subsection{Nombres de Ramsey}

On obtient des ingégalités de la forme \(L(k_1, ..., k_r, k_{r+1}, ..., k_{r+s}) - 2 \geqslant a \, (L(k_1, ..., k_r) - 2) + d\). Les "\(k_i\) ajoutés" désignent \( k_{r+1}, ..., k_{r+s}\). Dans le tableau ci-dessous, \(b\)-template désigne un \(b\)-template non exprimable en tant que template. Les anciennes valeurs viennent de \cite{rowleyramseysat}.
\begin{table}[H]\begin{center}
\begin{tabular}{| c | c | c | c | c |}
	\hline
	\(k_i\) ajoutés & ancien (a, d) & nouveau (a, d) & type de template & coloriage final \\
	\hline
	3, 3 & 9, 4 & 10, 2 & \(b\)-template & défaut \\
	\hline
	3, 4 & 18, 7 & 19, 2 & \(b\)-template & défaut \\
	\hline
	3, 5 & 30, 12 & 31, 3 & \(b\)-template & spécifique \\
	\hline
	4, 5 & 51, 21 & 55, 23 & template & défaut \\
	\hline
\end{tabular}
\end{center}\end{table}

En ce qui concerne les bornes inférieures, on a  \(R_8(3) \geqslant 5\,364\) et \(R_{13}(3)\geqslant 2\,038\,284\) (contre respectivement \(5\,288\) dans \cite{RowleyRamsey} et \(2\,011\,292\) dans \cite{schurboyz}), et plus généralement une amélioration pour les \(R_{5 k + 3}(3), \forall k \in \mathbb{N}^*\). On a aussi \(R(3 , 4 , 5 , 5) \geqslant 764\) (contre 729 dans \cite{rowleyramseysat}). Il existe de nombreuses inégalités récursives et constructions pour les nombres de Ramsey. Afin de savoir quelles bornes précisément ont été améliorées, il faudrait recenser chacunes des bornes inférieures et des méthodes existantes puis calculer toutes les bornes et voir si cetaines sont améliorées. Il est probable que les inégalités de cette section permettent d'améliorer les bornes inférieures pour d'autres nombres de Ramsey mais je n'ai pas effectué cette vérification.

@@

>>>>

Ramsey numbers and Schur numbers are defined in Section \ref{sec:def}. Section \ref{sec:temp} defines the 
\(b\)-templates and describes the corrresponding construction. Section \ref{sec:results} states and proves the main 
results on \(b\)-templates and on the construction. Section \ref{sec:bounds} records the new inequalities and some new 
lower bounds. The improved templates can be found in Appendix \ref{appendix}.

@@

Let \(a \in \mathbb{N}^*\), \(b \in [\![0, a - 1]\!]\) and \(r \in \mathbb{N}\). Let 
\((k_c)_{1 \leqslant c \leqslant r} \in {[\![3, + \infty [\![}^r\). Let \(T \in \mathcal{T}_b(k_1, ..., k_r, t; a)\) 
and denote by \(A_1, ..., A_{r+1}\) the associated partition. Let \(d \in [\![0, a]\!]\). Let \(F\) be a coloring of 
\([\![1, d]\!]\) with \(r +1\) colors and denote by \(B_1, ..., B_{r + 1}\) the associated partition. The coloring 
\(F\) is said to be compatible with the \(b\)-template \(T\) if it satisfies:
\begin{enumerate}[(i)]
\item \(a - b \notin B_{r + 1}\)
\item \(\forall (x,y) \in A_{r+1}^2, (x + y > a + b ~\text{and}~ \pi_{a, b}(x + y) \leqslant d) \implies \pi_{a, b}
	(x + y)  \notin B_{r+1}\)
\item \(\forall (x,y) \in A_{r+1} \times B_{r+1}, x + y \leqslant d \implies x + y  \notin B_{r+1}\)
\item \(\forall c \in [\![1, r]\!], \forall k \in [\![1, k_c-1]\!], \forall x \in \TS_{b, k}(A_c), \forall y \in 
	\mathfrak{I}_{k_c - k - 1}(B_c), \\
	\begin{array}{l}
		\mspace{30mu} 
		\left\{ \text{or} 
		\begin{array}{l}
			\exists (i, j) \in {[\![1, k]\!]}^2, \pi_{a, b}(x_j - x_i) \notin A_c, \\
			\exists (i, j) \in [\![1, k]\!] \times [\![1, k_c - k - 1]\!], \\
			\mspace{30mu}
				\left \{
				\begin{array}{l l}
					(y_j - x_i) \mod a \notin A_c & ~\text{if}~ (y_j - x_i) \mod a > b \\
					(y_j - x_i) \mod a \notin A_c ~\text{and}~ \pi_{a, b}(y_j - x_i) \notin A_c & ~\text{if}~ (y_j - x_i) \mod a 
						\leqslant b \\
				\end{array}
				\right. \\
			\exists (i, j) \in {[\![1, k_c - k - 1]\!]}^2, i < j ~\text{and}~ \pi_{a, b}(y_j - y_i) \notin A_c.
		\end{array}
		\right.
	\end{array}\)
\end{enumerate}
\end{definition}

@@

\begin{corollary}[Inégalités pour les nombres de Ramsey linéaires]
\label{thm:ineq}
Soit \(r \in \mathbb{N}\) et soit \((k_c)_{1 \leqslant c \leqslant r} \in {[\![3, +\infty[\![}^r\). Soit \(s \in \mathbb{N}\) et soit \((l_c)_{1 \leqslant c \leqslant s} \in {[\![3, +\infty[\![}^s\). Soient \(a \in \mathbb{N}^*\) et \(b \in \mathbb{N}\) tels que \(\mathcal{T}_b(l_1, ..., l_s, t; a) \neq \varnothing\), et on fixe \(T\) un tel \(b\)-template. Soit \(d \in [\![1, a]\!]\) tel qu'il existe un coloriage de \([\![1, d]\!]\) à \(s + 1\) couleurs compatible avec \(T\).
Alors \(L(k_1, ..., k_r, l_1, ..., l_s) - 2 \geqslant a (L(k_1, ..., k_r) - 2) + b + d\).
\end{corollary}

@@

\section{New inequalites and lower bounds}
\label{sec:bounds}

As in \cite{schurboyz} and \cite{rowleyramseysat}, the \(b\)-templates described in this article were found by encoding 
the existence of the templates into a boolean formula and then using a SAT solver (here, the parallel version of the 
Lingeling SAT solver \cite{Lingeling2017}).

\subsection{Multicolor Ramsey numbers}

The templates in Appendix \ref{appendix} provide inequalities of the form 
\(L(k_1, ..., k_r, k_{r+1}, ..., k_{r+s}) \geqslant a \, L(k_1, ..., k_r) - 2 + d\). In table \ref{tab:ineq},"added 
\(k_i\)'s" denotes the \( k_{r+1}, ..., k_{r+s}\) in the previous inequality and "\(b\)-template" indicates a 
\(b\)-template which cannot be expressed as a sf-template.

\begin{table}[H]
\begin{center}
\caption{Summarizing table of the new inequalities}
\begin{tabular}{| c | c | c | c | c |}
	\hline
	added \(k_i\)'s & former (a, d) & new (a, d) & template type & final coloring \\
	\hline
	3, 3 & 9, 4 \cite{AbbottHanson} & 10, 2 & \(b\)-template & default \\
	\hline
	3, 4 & 18, 7 \cite{rowleyramseysat} & 19, 2 & \(b\)-template & default \\
	\hline
	3, 5 & 30, 12 \cite{rowleyramseysat} & 31, 3 & \(b\)-template & specific \\
	\hline
	4, 5 & 51, 21 \cite{rowleyramseysat} & 55, 23 & sf-template & default \\
	\hline
\end{tabular}
\end{center}
\label{tab:ineq}
\end{table}

\begin{table}[H]
\begin{center}
\caption{Summarizing table of the new lower bounds for small Ramsey numbers}
\begin{tabular}{| c | c | c |}
	\hline
	Ramsey number & former lower bound & new lower bound \\
	\hline
	\(R_8(3)\) & \(5\,288\) \cite{RowleyRamsey} & \(5\,364\)  \\
	\hline
	\(R_{13}(3)\) & \(2\,011\,292\) \cite{schurboyz} & \(2\,038\,284\) \\
	\hline
	\(R(3 , 4 , 5 , 5)\) & 729 \cite{rowleyramseysat} & 764 \\
	\hline
\end{tabular}
\end{center}
\label{tab:bounds}
\end{table}
The template in \(\mathcal{L}_2(3, 3, t; 10)\) induces an inequality which improves the lower bounds for 
\(R_{5 n + 3}(3)\) for all \(n \in \mathbb{N}^*\).

\subsection{Schur numbers}

The existence of a 2-template with width 10 and 3 colors which cannot be expressed as a sf-template provides an improved 
inequality. The former corresponding inequality is \(S(n + 2) \geqslant 9 \, S(n) + 4\) \cite{AbbottHanson}.

\begin{proposition}
	\[\forall n \in \mathbb{N}, S(n + 2) \geqslant 10 \, S(n) + 2\]
\end{proposition}

This inequality provides the following improved lower bounds: \(S(8) \geqslant 5\,362\) and 
\(S(13) \geqslant 2\,038\,282\) (the former lower bounds were respectively \(5\,286\) \cite{RowleyRamsey} and 
\(2\,011\,290\) \cite{schurboyz}). More generally, this inequality improves the lower bounds for \(S(5 n + 3)\) for all 
\(n \in \mathbb{N}^*\).

\section{Acknowledgements}

This study results from a long-term project performed within the framework of the Project Cluster "Training for 
Research" of CentraleSupélec. I also would like to address warm thanks to Paul Casteras, Thibaut Pellerin and Yann 
Portella with whom I started working on Schur and weak Schur numbers.

@@

\appendix

\section{\(b\)-Templates}
\label{appendix}

\begin{table}[H]
\caption{\(T \in \mathcal{T}_2(3, 3, t; 10)\)}
\begin{center}
\begin{tabular}{c}
	\begin{tabular}{| c | c |}
		\hline
		1 & 1, 5, 8, 12 \\
		\hline
		2 & 2, 6, 7 \\
		\hline
		3 & 3, 4, 9, 10, 11 \\
		\hline
	\end{tabular} \\ \\
	 Final coloring: \(\varnothing\)
\end{tabular}
\end{center}
\end{table}

\begin{table}[H]
\caption{\(T \in \mathcal{T}_2(3, 4, t; 19)\)}
\begin{center}
\begin{tabular}{c}
	\begin{tabular}{| c | c |}
		\hline
		1 & 1, 6, 9, 13, 17, 21 \\
		\hline
		2 & 2, 7, 8, 10, 14, 15, 16 \\
		\hline
		3 & 3, 4, 5, 11, 12, 18, 19, 20 \\
		\hline
	\end{tabular} \\ \\
	 Final coloring: \(\varnothing\)
\end{tabular}
\end{center}
\end{table}

\begin{table}[H]
\caption{\(T \in \mathcal{T}_2(3, 5, t; 31)\)}
\begin{center}
\begin{tabular}{c}
	\begin{tabular}{| c | c |}
		\hline
		1 & 1, 5, 9, 12, 16, 19, 23, 26, 30 \\
		\hline
		2 & 2, 6, 7, 8, 10, 11, 17, 18, 24, 25, 27, 28, 29, 33 \\
		\hline
		3 & 3, 4, 13, 14, 15, 20, 21, 22, 31, 32 \\
		\hline
	\end{tabular} \\ \\
	 Final coloring: \(F(1) = 1\)
\end{tabular}
\end{center}
\end{table}

\begin{table}[H]
\caption{\(T \in \mathcal{T}_0(5, 4, t; 55)\)}
\begin{center}
\begin{tabular}{c}
	\begin{tabular}{| c | c |}
		\hline
		1 & 1, 5, 6, 7, 9, 10, 12, 14, 15, 17, 18, 19, 23, 25, 31, 48, 54 \\
		\hline
		2 & 2, 3, 4, 8, 11, 13, 16, 20, 21, 22, 26, 28, 51, 52, 53 \\
		\hline
		3 & 24, 27, 29, 30, 32, 33, 34, 35, 36, 37, 38, 39 \\
		 & 40, 41, 42, 43, 44, 45, 46, 47, 49, 50, 55 \\
		\hline
	\end{tabular} \\ \\
	 Final coloring: \(\coldef_0(T)\)
\end{tabular}
\end{center}
\end{table}
